\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-27 papers/jfp/7-experience.tex
Proofing and tweaking

As a set of authors, we've written a paper about the \Shake build system \cite{mitchell2012shake},
applying \Shake to building GHC \cite{hadrian}, and a paper about the categorisation of build
systems \cite{mokhov2018buildsystems}. The original motivation of the research in that final paper was to create
``Cloud Shake'', and use it for the GHC build system. In this section we tie these papers together,
reflecting on our experience since these papers.

@@

The original design of \Shake hasn't changed since the initial paper. Since that paper there have been roughly 5,000 commits to the \Shake project\footnote{\url{https://github.com/ndmitchell/shake}}. They add concepts like resources (if two rules content from a single external resource), rewrite serialisation to be faster, documentation including a website\footnote{\url{https://shakebuild.com}}, and add lots of tests. The biggest change in that time period was moving from blocking threads to continuations for the suspending scheduler. The most visible change is switching \hs{*>} for \hs{\%>}\footnote{A conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost all external and internal details remain the same.

We consider the lack of change to suggest that \Shake is based on fundamental principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}. There are two main aspects of the original \Shake paper \cite{mitchell2012shake} that are described more clearly in this paper, the first is the rebuilder, which is described as verifying step traces in \S\ref{sec-step-traces}. The second is the power of \Shake tasks (called actions in the original paper. Originally these tasks were described as being equivalent to:

\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v r = Finished r
                  | Depends v (v -> Action k r)
\end{minted}

In some sense this is the ``free'' version of the \hs{Task Monad} \cite{free-applicatives} -- the two types form an isomorphism. The originally paper uses concrete types \hs{Key} and \hs{Value}, we generalise these types to \hs{k} and \hs{v}, and also add \hs{r} so that \hs{Action k v} can be an instance of \hs{Monad}. We can write that \hs{Monad} instance, and use it to write conversion functions to/from our \hs{Task} type:

@@

Similary, \Make tasks were described as equivalent to

@@

Once again, this type is isomorphic to \hs{Task Applicative}, and we can define a similar \hs{Applicative} instance and conversion functions. By describing these types using \hs{Task} we are able to pinpoint the differences (\hs{Monad} vs \hs{Applicative}) and use existing literature to prove what is and isn't possible.

@@

Converting \Shake into \Cloud \Shake wasn't a difficult process once armed with a roadmap. The key was the introduction of two new functions:

@@

The integration of these functions into \Shake is also interesting. We found the most expedient design was to leave \Shake with a verifying trace, but if the verifying trace doesn't work, we use then consult the constructive trace. By bolting constructive traces onto the side of Shake we avoid reengineering of the central database. We haven't found any significant downsides from the bolt-on approach thus far, so it may be a sensible route to go even if developing from scratch -- allowing an optimised verified trace implementation in many cases, and falling back to a more complex implementation only rarely.

>>>>

As a set of authors, we've written a paper about the \Shake build system \cite{mitchell2012shake},
applying \Shake to building GHC \cite{hadrian}, and a paper about the categorisation of build
systems \cite{mokhov2018buildsystems}. The original motivation of the research in that final paper was to create
``\Cloud \Shake'', and apply it to the GHC build system. In this section we tie these papers together,
reflecting on our experience since these papers.

@@

The original design of \Shake hasn't changed since the initial paper, but the implementation has continued to mature -- there have been roughly 5,000 subsequent commits to the \Shake project\footnote{\url{https://github.com/ndmitchell/shake}}. These commits add concepts like resources (if two rules contend on a single external resource), rewriting serialisation to be faster, documentation including a website\footnote{\url{https://shakebuild.com}}, and add lots of tests. The biggest change in that time period was moving from blocking threads to continuations for the suspending scheduler. The most visible change is switching \hs{*>} to \hs{%>} for defining rules\footnote{A conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost all external and internal details remain the same.

We consider the lack of change suggestive that \Shake is based on fundamental principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}. There are two main aspects to the original \Shake paper \cite{mitchell2012shake} that are described more clearly in this paper, Firstly, the rebuilder can now be described as verifying step traces as per \S\ref{sec-step-traces}, with a much clearer relationship to the unoptimised verifying traces in \S\ref{sec-verifying-traces}. The second is the power of \Shake tasks (called actions in the original paper). Originally these tasks were described using the data type:

\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v r = Finished r
                  | Depends v (v -> Action k v r)
\end{minted}

In some sense, this is the ``free'' version of \hs{Task Monad} \cite{free-applicatives} -- the two types form an isomorphism. The original paper uses concrete types \hs{Key} and \hs{Value}, we generalise these types to \hs{k} and \hs{v}, and also add \hs{r} so that \hs{Action k v} can be an instance of \hs{Monad}. We define that \hs{Monad} instance, and use it to write conversion functions to/from our \hs{Task} type:

@@

\noindent Similarly, \Make tasks were described as equivalent to

@@

Once again, this type is isomorphic to \hs{Task Applicative}, and we can define a similar \hs{Applicative} instance and conversion functions. By describing these types using \hs{Task} we are able to describe the differences more concisely (\hs{Monad} vs \hs{Applicative}) and use existing literature to determine what is and isn't possible.

@@

Converting \Shake into \Cloud \Shake wasn't a difficult process once armed with a roadmap from \cite{mokhov2018buildsystems}. The key was the introduction of two new functions:

@@

The integrate these functions into \Shake we found the most expedient design was to leave \Shake with a verifying trace, but if the verifying trace doesn't skip building, we consult the constructive trace. By bolting constructive traces onto the side of Shake we avoid reengineering of the central database. We haven't found any significant downsides from the bolt-on approach thus far, so it may be a sensible route to go even if developing from scratch -- allowing an optimised verified trace implementation in many cases, and falling back to a more complex implementation only rarely.
